An Interactive Tutorial on Hash Array Mapped Tries

### 2024-08-23 paper/paper.tex
revise

\begin{abstract}
Clojure provides a suite of persistent data structures
implemented by Hickey based on previous work by Bagwell.
In this tutorial, we use a ported implementation of
Hickey's Java implementation to Clojure to learn
how Hash Array Mapped Tries work.
\end{abstract}

@@

Hash Array Mapped Tries (HAMT) have rocked the functional programming
world with a fast, immutable and persistent alternative
to a hash map.
First described by Bagwell~\cite{bagwell2001ideal},
they are featured in mainstream functional programming
languages like Clojure and Scala, and have been ported
to many others.

@@

To understand with a hash array mapped trie is, 
we first give some definitions.
A \textit{trie} is a way of formatting key/value pairs
in a tree, where values are leaves and keys are spread
across the paths to those nodes.
Key prefixes occur on the shallow levels of the tree,
and suffixes occur closer to the leaves.
A \textit{bit trie} assumes the mapping keys are strings
of bits. Each level consumes one or more bits to index
its elements.
An \textit{Array Mapped Trie}
maps the bits of array indices as a bit trie.

In this paper, we explore Clojure's persistent hash
map implementation
(Section~\ref{clojure-hamt}).
It was implemented by Hickey~\cite{hickey2008clojure}, extending
Bagwell's original formulation~\cite{bagwell2001ideal}
to be persistent.
Persistent data structures use \textit{structural sharing}
when extending themselves, so Clojure necessarily enforces
hash maps to be \textit{immutable}.

@@

\begin{enumerate}
  \item We walkthrough the mechanics behind HAMTs.
  \item We describe the internals of Clojure's persistent HAMT implementation.
  \item We present a port of Clojure's HAMT from Java to Clojure
    for pedagogical purposes,
    featuring trie visualization (Figure~\label{port-visualize}).
  %\item We evaluate a Clojure-specific optimization for Clojure's HAMT.
  %\item We speculate on future directions for HAMT's.
\end{enumerate}

@@

Firstly, a HAMT represents a search tree
based on the \textit{hash} of its keys.
Each key is associated with a value.
Figure \ref{hashes} gives sample 32-bit hashes for six keys,
which we will use only in this section.

@@

\paragraph{Insertion}
Let's insert a mapping into an empty tree.
Since the tree is empty, we create the
first (root) level, level 0. This corresponds
to the first 5 bits of the hash. The maximum
branching factor is $2^5=32$, but since we
only need one entry, we create a \textit{resizable}
root node.

@@

\paragraph{Full nodes}
Once a resizable array reaches a certain threshold
size, it is no longer efficient to allocate
an array of length $2n$ to hold $n$ nodes.
For example, if we have over 16 entries, which
would require copying arrays over length 32, we could
instead once-and-for-all allocate a length 32
array where each member is a subtrie
(without a $\times$ flag)---we call this a \textit{full}
node.

@@

\paragraph{Hash collision nodes}
If two different keys hash to the same value,
we use a \textit{hash collision node}
to differentiate them. One approach
is to default to a linear search---with
the assumption that the hash function
has a low probabilities of collisions,
this should rarely be an issue.

@@

The return value can then 
be used bit \textit{and}ed
with the bitmap to return the value of the
desired bit in the bitmap.

@@

To retrieve the next array index, we count the number of 1's
below the given bit in the bitmap (assuming the given
bit is set to 1).
This number $i$ is the number of nodes \textit{before}
the node of interest---thus indexes $2i$ and $2i+1$
contain the key and value of interest.
To demonstrate this, say we have a bitmap
with the first 8 bits $1011\ 0010$, and
otherwise zeros.
Since four bits are 1, we have four nodes
in an array of size $4\cdot2=8$.

@@

In the implementation, \texttt{bit}
is a 32-bit integer with exactly 1 bit set to 1.
By decrementing \texttt{bit}, we aquire
a useful mask to isolate all the necessary
bits in the bitmap.
For example, if bitmap was \texttt{1000}---that
is, isolating the 4th bit---decrementing it results 
in \texttt{0111}.
Bit \textit{and}ing \texttt{0111}
with bitmap then isolates the 1st-3rd bits, which
we can then use to count the number of 1's
below \texttt{bit} in \texttt{bitmap}.

@@

Clojure's implementation of HAMT is implemented on the JVM,
which only has signed 32-bit integers.
The HAMT implementation, however, treats hashes as
arbitrary strings of 32-bits, so we need to emulate
unsigned arithmetic operations.

@@

The difference is, \texttt{>>} preserves the most significant
bit, while \texttt{>>>} replaces it with 0.
%
\begin{verbatim}
1000 1101 >>  1 = 1100 0110    //signed
1000 1101 >>> 1 = 0100 0110    //unsigned
\end{verbatim}
%
We always want \textit{unsigned} bit operations, because no bits
are special in a hash, or in a bitmap.

>>>>

\begin{abstract}
Clojure provides a suite of persistent data structures
implemented by Hickey based on previous work by Bagwell.
This tutorial teaches
how Hash Array Mapped Tries work
using a Clojure port of
Hickey's Java implementation.
\end{abstract}

@@

Hash Array Mapped Tries (HAMT) provide a key tool for functional programmers
as fast, immutable, and persistent alternatives to hash maps.
First described by Bagwell~\cite{bagwell2001ideal},
they are featured in mainstream functional programming
languages like Clojure and Scala, and have been ported
to many others.

@@

To understand with a hash array mapped trie is, 
we first give some definitions.
A trie is a way of formatting key/value pairs
in a tree, where values are leaves and keys are spread
across the paths to those nodes.
Key prefixes occur on the shallow levels of the tree,
and suffixes occur closer to the leaves.
A bit trie assumes the mapping keys are strings
of bits. Each level consumes one or more bits to index
its elements.
An Array Mapped Trie
maps the bits of array indices as a bit trie.

In this paper, we explore Clojure's persistent hash
map implementation
(Section~\ref{clojure-hamt}).
It was implemented by Hickey~\cite{hickey2008clojure}, extending
Bagwell's original formulation~\cite{bagwell2001ideal}
to be persistent.
Persistent data structures are extended using structural sharing,
so Clojure necessarily enforces hash maps to be immutable.

@@

\begin{enumerate}
  \item We walk through the mechanics behind HAMTs.
  \item We describe the internals of Clojure's persistent HAMT implementation.
  \item We present a port of Clojure's HAMT from Java to Clojure
    for pedagogical purposes,
    featuring trie visualization (Figure~\label{port-visualize}).
  %\item We evaluate a Clojure-specific optimization for Clojure's HAMT.
  %\item We speculate on future directions for HAMT's.
\end{enumerate}

@@

Firstly, a HAMT represents a search tree
based on the hash of its keys.
Each key is associated with a value.
Figure \ref{hashes} gives sample 32-bit hashes for six keys,
which we will use only in this section.

@@

\paragraph{Insertion}
Let's insert a mapping into an empty tree.
Since the tree is empty, we create the
first (root) level, level 0. This corresponds
to the first 5 bits of the hash. The maximum
branching factor is $2^5=32$, but since we
only need one entry, we create a resizable
root node.

@@

\paragraph{Full nodes}
Once a resizable array reaches a certain threshold
size, it is no longer efficient to allocate
an array of length $2n$ to hold $n$ nodes.
For example, if we have over 16 entries, which
would require copying arrays over length 32, we could
instead once-and-for-all allocate a length 32
array where each member is a subtrie
(without a $\times$ flag)---we call this a full
node.

@@

\paragraph{Hash collision nodes}
If two different keys hash to the same value,
we use a hash collision node
to differentiate them. One approach
is to default to a linear search---with
the assumption that the hash function
has a low probabilities of collisions,
this should rarely be an issue.

@@

The return value can then 
be used, combined with the bitmap using bitwise AND,
to return the value of the desired bit in the bitmap.

@@

To retrieve the next array index, we count the number of 1's
below the given bit in the bitmap (assuming the given
bit is set to 1).
This number $i$ is the number of nodes before
the node of interest---thus indexes $2i$ and $2i+1$
contain the key and value of interest.
To demonstrate this, say we have a bitmap
with the first 8 bits $1011\ 0010$, and
otherwise zeros.
Since four bits are 1, we have four nodes
in an array of size $4\cdot2=8$.

@@

In the implementation, \texttt{bit}
is a 32-bit integer with exactly 1 bit set to 1.
By decrementing \texttt{bit}, we aquire
a useful mask to isolate all the necessary
bits in the bitmap.
For example, if bitmap was \texttt{1000}---that
is, isolating the 4th bit---decrementing it results 
in \texttt{0111}.
Combining \texttt{0111}
and the bitmap with bitwise AND then isolates the 1st-3rd bits, which
we can then use to count the number of 1's
below \texttt{bit} in \texttt{bitmap}.

@@

Clojure's implementation of HAMT is implemented on the JVM,
where 32-bit integers are signed.
The HAMT implementation, however, treats hashes as
arbitrary strings of 32-bits, so we need to emulate
unsigned arithmetic operations.

@@

The difference is, \texttt{>>} preserves the most significant
bit, while \texttt{>>>} replaces it with 0.
%
\begin{verbatim}
1000 1101 >>  1 = 1100 0110    //signed
1000 1101 >>> 1 = 0100 0110    //unsigned
\end{verbatim}
%
Unsigned bit operations are necessary because no bits
are special in hashes and bitmaps.
